Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-03-24 agujournaltemplate.tex
Discussion 98% done - pre save

    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

@@

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
\label{fig:divts}
\end{figure}

@@

\begin{table}
\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\label{table:clustering}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & ***\\
u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
Isotherm \qty{21}{\celcius} & 5293.0 & 0.034 & **\\
\addlinespace
Temperature & 4763.0 & 0.003 & ***\\
$NO_3$ & 6378.0 & 0.018 & **\\
$PO_4$ & 7359.0 & 0.012 & **\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
Salinity & 7636.0 & 0.002 & ***\\
\addlinespace
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & **\\
\bottomrule
\end{tabular}
\end{table}

@@

However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}. We can see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!

@@

When comparing the community between the two clusters...
- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column. When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
- Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of the fluorometric chlorophyll reduction between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. 
- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy

In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}. In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
Alterations in physical mixing affected phytoplankton dynamics,

- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
- Discuss how similar Part 1 and 2 of Cluster 1 are.. Why are they clustered together?
- Discuss why we do not observe a clear distinction in Diversity between the two clusters, but mention the species turnover difference and the community evenness increase.

@@

When comparing the environmental conditions between the two clusters...
- When looking at distributions of measured environmental variables, there is a clear trend signifying the reduction in upwelling conditions during the period from 2004 to 2014 (cluster 2).
- We have lower wind speed, lower Isotherm, higher Temp, lower Nuts (except Silicate -> river runoff) lower salinity -> less mixing, higher precipitation.

>>>>

    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024). 
    %The term "Atlantic Multidecadal Oscillation" (AMO) describes the large-scale fluctuations in Atlantic sea surface temperature (SST) that occur over multiple decades \cite{kerr2000north}. This designation is intended to underscore the distinct multidecadal nature of these oceanic variations in contrast to the inter-annual variability linked with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}.

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

@@

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\begin{center}
\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals. (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling. Transparent dots show monthly raw values. (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
\label{fig:divts}
\end{figure}

@@

\begin{table}
\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters. Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
\label{table:clustering}
\centering
\begin{tabular}[t]{lrrl}
\toprule
Variable & W & p.value & Significance\\
\midrule
AMO & 3494.0 & 1.86e-14 & ***\\
MEI v.2 & 9523.5 & 0.006 & ***\\
u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & **\\
\addlinespace
Temperature & 4763.0 & 0.003 & ***\\
$NO_3$ & 6378.0 & 0.018 & **\\
$PO_4$ & 7359.0 & 0.012 & **\\
$SiO_4$ & 5998.0 & 0.736 & n.s.\\
Salinity & 7636.0 & 0.002 & ***\\
\addlinespace
Genus Richness & 5913.0 & 0.954 & n.s.\\
Shannon Index & 5142.0 & 0.108 & n.s.\\
Pielou Index & 4874.0 & 0.029 & **\\
\bottomrule
\end{tabular}
\end{table}

@@

However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}. We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
To identify the variables that can best predict changes in the phytoplankton community, we used a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices and in situ measurements of bottom-up drivers as predictors. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis found that the strongest predictors of changes within the community are not in situ measurements, but actually the AMO and the MEI v.2 climate indices. These are followed in importance by nitrate concentration and temperature in the upper \qty{100}{meters} (see Figure \ref{fig:GF}).

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that actually separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2. This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1. Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.

Using microscopy cell counts as a basis for this analysis meant that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitation of identifying smaller species via light microscopy. Grouping cell counts by genera does create a loss of resolution, but is based on the assumption that species within a genus generally show similar ecophysiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, but also some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$). Within these groups, we see differing threshold responses in cumulative importance along the predictor ranges.

AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south. This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns. Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies. This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin. The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.

The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag. We also tested a 12 month lag and found no significant improvement.
The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions. For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events. Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range. Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly all found in Cluster 1 (see Figure \ref{fig:clustering} $e$).

The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, temperature and salinity, in that order. Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}. For dinoflagelltes and diatoms we see a strong threshold response around \qty{7}{\micro\molar}, which would indicate that strong upwelling conditions drive blooms of these functional groups. The "other" functional group including \textit{Mesodinium} and \textit{Eutreptria} seems to have a threshold response at even higher nitrate concentrations (around \qty{9}{\micro\molar}), indicating an effect on their abundance by very strong upwelling. Similarly, for temperature dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper\qty{100}{m} driven by strong upwelling. For the other predictors, we do not see strong threshold responses.

Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of microphytoplankton community changes in the Cariaco basin, in our limited set of bottom-up drivers. This is an interesting result, as it would be expected that more direct measures of ecosystem variability could better predict shifts in abundances. One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month. Climate indices are generally detrended and integrated measurements of long-term changes, much more matching in scale to the dynamics of long-term changes within a phytoplankton community. Yet we do not see such strong predictive power in the ERA5 climate reanalysis-derived wind speed, which also aggregates measurements and model output over an entire month. 
Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$).

Our limitiation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affects phytoplankton community structure. Due to the lack of data resolution for higher trophic levels, and the lack of zooplankton data for the start of the time series, this might require mechanistic modeling studies to investigate this aspect of the CARIACO time series further.
Also the limitations of microscopy.

Comparison to Sargasso Sea, changes in temp and upwelling and community, BUT export remains steady \cite{lomas_increased_2010, lomas_adaptive_2022}. Most likely mediated and strong influence of grazers, not resolved in our study.

- Mention other papers correlatting AMO to ecosystems!

Generally we see a strong impact of decadal-scale climatic oscilaltions on shifts in the tropical coastal ecosystem of the Cariaco basin. The microphytoplankton community recovered to previous levels, despite general warming trends, highlighting the resilience of ocean ecosystems under global change.

@@

@@
